A Note on Convergence Diagnostics for Multiple Imputation using Chained Equations (MICE)

### 2019-12-11 Simulation/Writeup/NoteConvergence.Rnw
ran full simulation with convergence diagnostics per var instaed of max

%% - \author{} with primary affiliation
%% - \Plainauthor{} without affiliations
%% - Separate authors by \And or \AND (in \author) or by comma (in \Plainauthor).
%% - \AND starts a new line, \And does not.
\author{Hanne Oberman\\Utrecht University}
   % \And Second Author\\Plus Affiliation}
\Plainauthor{Hanne Oberman}

@@

%% - \Abstract{} almost as usual
\Abstract{
  This Research Report contains a simulation study that serves as the basis of the technical paper that will be submitted for publication in \emph{Journal of Statistical Software}. I have chosen to use the first format from the Research Report guidelines: ``\emph{It is written as a (mini) thesis, with an introduction, methods section, some results (i.e., preliminary analyses, or pilot simulations), and a discussion of results. The length of the research report should be maximally 2500 words of text (without references list and or tables and figures). Please do not include appendices, and no more than 6 tables or figures. Table and Figure captions do not count towards the word limit. An abstract may be included, but is not necessary}''. I aim to publish a pre-print of this research report on \href{https://arxiv.org/}{ArXiv}. That way, I can refer to the simulation study described here, without 'bulking up' the technical paper that I want to submit for publication in JSS. Another option would be to attach this note as online appendix to the technical paper on ShinyMICE. \textbf{TO ADD: appendix with extra results tables with different ampute and or very high or very low correlations between predictors}.
%"It is the first half of the thesis, i.e., there are no results included yet, but the report contains a full introduction including a literature review and a methods section that contains details about the data, instruments and or statistical procedures". The goal was to develop novel methodology and guidelines for evaluating multiple imputation methods, and implement these in an interactive evaluation framework for multiple imputation: \pkg{ShinyMICE}.
  }

@@

\textbf{ADD some references in this paragraph!} To obtain valid inferences with MI, the variability between imputations should be properly represented. If the variance between imputations is under-estimated, confidence intervals around estimates will be too narrow, which can yield spurious results. On the other hand, over-estimation of the variance between imputations results in unnecessarily wide confidence intervals, which can be costly because it lowers the statistical power. Since both of these situations are undesirable, imputations and their variability should be properly evaluated. Such evaluation measures are currently missing or under-developed in MI software, like the world-leading \pkg{mice} package \citep{mice} in \proglang{R} \citep{R}. 
%Therefore, I will answer the following question: 'Which measures are vital for evaluating the validity of multiply imputed data?'.
The aim of this research project is to investigate the convergence properties of the MI algorithm that is implemented in \pkg{mice}: `Multiple Imputation using Chained Equations' (MICE). This note addresses the question: `How to diagnose convergence of the multiple imputation algorithm implemented in \pkg{mice}?'.

@@

The convergence guidelines introduced in this paper are developed to be integrated into the \pkg{mice} environment \citep{mice} in \proglang{R} \citep{R}. This note therefore follows notation and conventions of \cite{mice}. Deviations from the `original' notation by \cite{rubin87} are described in \citep[\S~XYZ]{buur18}.

@@

This note focusses on the algorithmic properties of the imputation step. The algorithm employed within this step has an iterative nature. That is, before drawing $m$ imputed values for each missing data point in $Y_{mis}$, a `chain' of potential values is considered. Each of the $m$ chains starts with an initial value, drawn randomly from $Y_{obs}$. The chains are terminated after a predefined number of iterations. Only the ultimate value that a chain lands on is imputed, and hence used in the next steps (analysis and pooling). The collection of values (or summary statistics) across iterations of each of the $m$ imputations will be referred to as an `imputation chain'.

@@

The methodology for diagnosing convergence of multiple imputation algorithms has not been systematically studied \cite{buur18}. Yet, reference manuals of promiment MI software do mention the importance of evaluating algorithmic convergence (e.g., \cite[p.XYZ]{mice}, XYZ, XYZ). This section provides a brief review of how convergence is defined in the context of iterative algorithmic procedures, what diagnostic tools are available, and which of these may be applicable to MI algorithms.

@@

\textbf{What is convergence in general?}

We can never be certain of convergence, therefore convergence diagnostics evaluate signs of non-convergence \citep{hoff09}. % In general: "In general, you cannot know for sure if your chain has converged. But sometimes you can know if your chain has not converged, so we at least check for this latter possibility" \cite[p.~101]{hoff09}
Convergence has two components \textbf{[look up source!!!]}: mixing between chains and stability over iterations within chains. Ideally, both components will be satisfied: there are no trends within chains, and the chains intermingle such that the only difference between the chains is caused by the randomness induced by the algorithm. Diagnose non-convergence in the mixing sense: $\widehat{R}$. Diagnose non-convergence in the stability sense: auto-correlation (AC). Also, it is possible to compute MC errors, or the Geweke statistic \citep{XYZ}. All of these methods evaluate the convergence of univariate scalars of interest (e.g., chain mean or chain variance). The convergence of multivariable statistics (i.e. relations between such scalars) may be evaluated via the eigenvector decomposition method proposed by McKay \citep{XYZ}. These three methods, however, are outside of the scope of this study. Here, the focus is mainly on $\widehat{R}$.

@@

We can apply $\widehat{R}$ to the mean (or to the first two moments) of the variables of interest. $\widehat{R}$ is defined by \cite{veth19} as follows: % The scalar summary of interest in this study is the average of the imputed values per variable. %As recommended by \cite{veht19} $\widehat{R}$ is computed as .... to be able to detect ... in the tails of the distribution. \cite[p.~5]{veth19}

@@

The second component of convergence ... interpreted as stability over iterations, or non-recurring. Non-recurrence can be evaluated with auto-correlation. Auto-correlation shows how dependent subsequent draws of an imputation chain are on the previous value. If there is a lot of dependence, draws at e.g. iteration five are significantly correlated with the value of the first draw. A high auto-correlation indicates dependence within chains. The magnitude of the AC can be interpreted qualitatively, or compared to the critical value of a \textbf{XYZ} distribution with \textbf{XYZ} degrees of freedom.

\textbf{Convergence of MI}

The convergence properties of multiple imputation algorithms lack scientific consensus \citep{taka17}. A review by \citep{murr18}  and some default MICE techniques might not converge to stable distributions at all \citep{murr18}. Convergence diagnostics for MI methods have not been systematically studied \citep{buur18}.  Moreover, 
A fundamental assumption of MICE is convergence of the algorithm. MICE is a type of Markov chain Monte Carlo (MCMC) algorithm. In general, the validity of inference resulting from MCMC algorithms is threatened by non-convergence. Hence we use convergence diagnostics to flag non-convergence. But it is not known whether conventional convergence diagnostics for MCMC methods work on MI data.

@@

\textbf{How is convergence checked now (current practice)?}

Conventional thresholds to diagnose non-convergence-- e.g., Gelman and Rubin's \citeyear{gelm92} statistic $\widehat{R} < 1.1$ --are not applicable on multiply imputed data \citep{lace07} \textbf{[Expand or remove Rhat here! Refer to future section?]}. Therefore, empirical researchers have to rely on visual inspection procedures that are theoretically equivalent to $\widehat{R}$ \citep{whit11} \textbf{[explain why this can still be appropriate, and numerical is not]}. Visually assessing convergence is not only challenging to the untrained eye, it might also be futile.

@@

\subsection{Data}
The simulated dataset is a finite population of $N=1000$. The data are simulated to solve a multiple linear regression complete data problem. The quantity of scientific interest is the estimated regression coefficient of predictor $X$ on outcome variable $Y$. The linear regression model is $Y \sim \beta_1 X + \beta_2 Z_1 + \beta_3 Z_2$, where $Y$ is the dependent variable, $X$ is an independent variable, and $Z_1$ and $Z_2$ are covariates. The data generating model of the predictors is a multivariate normal distribution with means structure $\mu$, and variance-covariance matrix $\Sigma$.

@@

\subsection{Simulation diagnostics}
The simulation diagnostics are as recommended by \cite{buur18}. These comprise of average bias, average confidence interval width, and empirical coverage rate (coverage probability) across simulations. %"We expect that in accordance with Neyman (1934), at least 95 percent of the confidence intervals should contain the true population value, although some room for simulation error should be taken into account to counteract the finite nature of our simulations"\cite{schou18}.
We could also look at distributional characteristics, and plausibility of imputed values, see \cite{vinknd} (n.d.). For now, this is outside of the scope of this study. % Or use: Then evaluate the regular diagnostics of MI simulations, see Vink, n.d., and  convergence. Diagnostics include absolute bias of the estimated regression coefficient, confidence interval width, and empirical coverage rate across simulations. Convergence is evaluated with $\widehat{R}$, or potential scale reduction factor.

@@

Figures \ref{fig:conv} and \ref{fig:sim} display results of the simulation study aggregated over 1000 repetitions. A subset of simulation conditions is presented in Table \ref{tab:results}. This section describes convergence diagnostics ($\widehat{R}$ and auto-correlation) and simulation diagnostics (bias, confidence interval width and coverage rate) separately.

@@

\textbf{$\widehat{R}$}. Steep decline up-to iteration twenty, then more gradual decrease up-to iteration forty, more or less stable after that. Conventional threshold $\widehat{R} < 1.2$ reached after seven or eight iterations, $\widehat{R} < 1.1$ after fourteen iterations for chain means and sixteen for chain variances, $\widehat{R} < 1.01$ not reached within the 100 iterations considered in this simulation study (see online appendix for the full table of results). \textbf{[REWRITE AS STORY, NOT JUST DATA.]}

\textbf{Auto-correlation.} Within the 100 iterations considered in this study, the AC decreases steeply, before a slow increase. The lowest AC observed is \Sexpr{round(min(dat[-1, c("AC chain mean", "AC chain var")]), 3)} at four iterations. There appears to be a trend towards zero from four iterations upwards. Auto-correlation does not plateau completely. The average absolute maximum ACs are negative up to 77 iterations for chain means, and for chain variances all iterations considered here have negative ACs.

@@

%% --------------------------------
\subsection{Simulation diagnostics}
\textbf{Bias.} From two iterations upwards, the average bias across repetitions is stable. Across iterations two to one hundred, bias fluctuates within a narrow range around zero. 
% Average bias is not zero because there is only one incomplete dataset to be imputed. Bias is a sample effect.

\textbf{Confidence interval width.} CIW does not show a clear trend across iterations. One might argue that there is some downward trending up-to thirthy to forty iterations. %After about twenty-five iterations the average CIW across all simulation conditions and repetitions is reached \Sexpr{round(mean(dat[, "CI width"]), 3)}.

\textbf{Coverage rate.} The coverage rate is more or less stable from two iterations upwards. On average, the coverage rate is somewhat higher than the expected nominal coverage of 95\% \cite{neym34}, namely \Sexpr{100*round(mean(dat[, "Cov. rate"]), 3)}\%.

\textbf{add loess line to plots to see when it becomes a flat line. also, give some implications for these measures.}

@@

@@

\textbf{Future research.} We would like to have convergence measures for multivariable statistics (scalars?) of interest. This is, however, dependent on the complete data model. The eigenvector decomposition method proposed by McKay (?) should be implemented. I could not find any resources to apply this method and it is outside the scope of this thesis to investigate how this approach could be implemented. \textbf{[this paragraph is too negative, explain scope in intro instead.]}

@@

\textbf{LEAVING THIS HERE FOR REFERENCE, PLEASE REMOVE BEFORE HANDING IN}

<<fig = TRUE, echo = FALSE>>=
plot.ts(dat[c(5:8, 2:4)], main = "", xlab = "Number of iterations")
@

>>>>

%% - \author{} with primary affiliation
%% - \Plainauthor{} without affiliations
%% - Separate authors by \And or \AND (in \author) or by comma (in \Plainauthor).
%% - \AND starts a new line, \And does not.
\author{\\Research Report (Preparation for Research Master Thesis, Research Seminar) \\ Methodology and Statistics for the Behavioural, Biomedical and Social Sciences \AND Hanne Oberman\\Utrecht University}
   % \And Second Author\\Plus Affiliation}
\Plainauthor{Hanne Oberman}

@@

%% - \Abstract{} almost as usual
\Abstract{
The ubiquitous problem of missing information may be be circumvented by implementing the framework of multiple imputation (MI). With MI, missing data points are `filled in' (i.e. imputed) several times, and statistical inference is performed on each imputed data set. The validity of inferences may be threathened by non-convergence of this iterative algorithmic procedure, but diagnostics to evaluate algorithmic convergence have not been implemented in MI software. This note investigates the performance of two convergence diagnostics ($\widehat{R}$ \citep{gelm92}, auto-correlation XYZ) by means of model-based simulation. To investigate convergence of the algorithm, the number of iterations is varied between one and one-hundred. Each of these 100 simulation conditions is replicated 1000 times. The convergence diagnostics are evaluated against three recommended simulation diagnostics (bias, confidence interval width, and coverage rate \citep{buur18}), since no benchmark diagnostic is available. This simulation study shows that three iterations may be sufficient to obtain ... \textbf{This Research Report contains a simulation study that serves as the basis of the technical paper that will be submitted for publication in \emph{Journal of Statistical Software}. I aim to publish a pre-print of this research report on \href{https://arxiv.org/}{arXiv}. That way, I can refer to the simulation study described here, without `bulking up' the technical paper that I want to submit for publication in JSS. Another option would be to attach this note as online appendix to the technical paper on ShinyMICE.}  
% This Research Report contains a simulation study that serves as the basis of the technical paper that will be submitted for publication in \emph{Journal of Statistical Software}. I have chosen to use the first format from the Research Report guidelines: ``\emph{It is written as a (mini) thesis, with an introduction, methods section, some results (i.e., preliminary analyses, or pilot simulations), and a discussion of results. The length of the research report should be maximally 2500 words of text (without references list and or tables and figures). Please do not include appendices, and no more than 6 tables or figures. Table and Figure captions do not count towards the word limit. An abstract may be included, but is not necessary}''. I aim to publish a pre-print of this research report on \href{https://arxiv.org/}{ArXiv}. That way, I can refer to the simulation study described here, without 'bulking up' the technical paper that I want to submit for publication in JSS. Another option would be to attach this note as online appendix to the technical paper on ShinyMICE.
  }

@@

To obtain valid inferences with MI, the variability between imputations should be properly represented \citep{rubin87, buur18}. If the variance between imputations is under-estimated, confidence intervals around estimates will be too narrow, which can yield spurious results. On the other hand, over-estimation of the variance between imputations results in unnecessarily wide confidence intervals, which can be costly because it lowers the statistical power. Since both of these situations are undesirable, imputations and their variability should be evaluated. Evaluation measures, however, are currently missing or under-developed in MI software, like the world-leading \pkg{mice} package \citep{mice} in \proglang{R} \citep{R}. %\textbf{ADD sentence to link eval measures to convergence!}
%Therefore, I will answer the following question: 'Which measures are vital for evaluating the validity of multiply imputed data?'.

The goal of this research project is to develop novel methodology and guidelines for evaluating MI methods. These evaluation measures and guidelines will subsequently be implemented in an interactive evaluation framework for multiple imputation to aid applied researchers in drawing valid inference from incomplete datasets. This note provides the theoretical foundation for one of these evaluation measures that is vital in evaluting MI procedures: a diagnostic to assess convergence of the MI algorithm. More specifically, this note focusses on the MI algorithm that is implemented in the \proglang{R} package \pkg{mice}: `Multiple Imputation using Chained Equations' (MICE). The reseach question that will be adressed is: %. a measure that is vital in evaluating one vital measure for evaluating MI procedures--an algorithmic convergence diagnostic. This note specifically investigates one vital evaluation measure: how to diagnose convergence of a multiple imputation algorithm. 
%This note focusses on one measure that is vital for evaluating the validity of multiply imputed data: a convergence diagnostic. of the MI algorithm. 
%Therefore, the aim of this research project is to investigate which measures are vital for . how the convergence properties of the MI algorithm that is implemented in \pkg{mice}: `Multiple Imputation using Chained Equations' (MICE). This note addresses the question: 
\emph{`How to diagnose convergence of the multiple imputation algorithm MICE?'.} \textbf{REMOVE THE :'s HERE!}

@@

The convergence guidelines introduced in this paper are developed to be integrated into the \pkg{mice} environment \citep{mice} in \proglang{R} \citep{R}. This note therefore follows notation and conventions of \cite{mice}. Deviations from the `original' notation by \cite{rubin87} are described in \citep[\S~2.2.3]{buur18}.

@@

This note focusses on the algorithmic properties of the imputation step. The algorithm employed within this step has an iterative nature. That is, before drawing $m$ imputed values for each missing data point in $Y_{mis}$, a `chain' of potential values is sampled. Each of the $m$ chains starts with an initial value, drawn randomly from $Y_{obs}$. The chains are terminated after a predefined number of iterations. Only the ultimate sample that a chain lands on is imputed, and subsequently used in the next steps (analysis and pooling). The collection of samples across iterations (between the initial value and the imputed value) for each of the $m$ imputations will be referred to as an `imputation chain'. The number of iterations will be denoted with $t$.

@@

The multiple imputation algorithm `MICE' is a special case of Markov chain Monte Carlo (MCMC) methods--the general framework of iterative algorithmic procedures. While methodology for diagnosing convergence of multiple imputation algorithms has not been systematically studied \citep{buur18}, several convergence diagnostics have been proposed for other types of MCMC methods (e.g., Gibbs samplers), or the framework in general.
% For the framework of itrIt is not known whether convergence diagnostics that are available for the general framework that MICE is a special case of are appropriate for the evaluation of MI data. This is the framework of Markoc chain Monte Carlo %Yet, reference manuals of promiment MI software do mention the importance of evaluating algorithmic convergence (e.g., \cite[p.XYZ]{mice}, \textbf{XYZ, XYZ}). 
This section provides a brief review of: 1) how convergence is defined in the general context of iterative algorithmic procedures; 2) what diagnostic tools are available; and 3) which of these may be applicable to MI algorithms.

Losely speaking, the aim of MCMC algorithms (e.g., Gibbs samplers, or MICE) is to sample values from an unknown target distribution (e.g., the posterior distribution of a parameter in Bayesian inference, or a distribution of missing values, respectively). Convergence is reached once the algorithm samples exclusively from the target distribution \citep{cowl96}. Since this target distribution is unknown by definition, convergence can only be monitored by evaluating signs of non-convergence \citep{hoff09}. Therefore, in practice, the definition of convergence consists of two components: `mixing' of chains together, and `stationarity' of individual chains \citep[p.~284]{gelm13}. The mixing component is satisfied when chains intermingle such that the only difference between the chains is caused by the randomness induced by the algorithm. Stationarity refers to the absence of trending within chains. \textbf{Figure XYZ shows a pathological case of non-convergence with respect to both components.}  
%The aim of MCMC procedures is to sample from an unknown (theoretical) distribution. The stability of the solution denpends on two components of convergence: should have converged, two components should be satisfied.

% Assessing convergence is difficult. We can never be certain of convergence, therefore convergence diagnostics evaluate signs of non-convergence \citep{hoff09}. % In general: "In general, you cannot know for sure if your chain has converged. But sometimes you can know if your chain has not converged, so we at least check for this latter possibility" \cite[p.~101]{hoff09}
% More importantly, in MCMC algorithms like MICE, "what is produced by the algorithm at convergence is not a single number or even a distribution, but rather a sample from a distribution" \citep[p.~883]{cowl96}. "automated convergence monitoring (as by a machine) is unsafe and should be avoided" \citep[p.~902]{cowl96}.
%Convergence has two components: `mixing' and `stationarity' \citep[p.~284]{gelm13}. Mixing refers to the variation between chains of samples. Stationarity has to do with stability of samples within chains. Ideally, both components will be satisfied: there are no trends within chains, and the chains intermingle such that the only difference between the chains is caused by the randomness induced by the algorithm.

The two components of convergence may be evaluated with convergence diagnostics. Diagnostics that target the stationarity component are auto-correlation \citep{gelm13}, numeric standard error \citep[which is commonly called `MC error']{gewe92}, and Raftery and Lewis's \citeyear{raft92} procedure to determine the increase in the necessary number of iterations due to dependence within chains. The mixing component of convergence, on the other hand, is primarily assessed with the `Gelman-Rubin statistic'--potential scale reduction factor $\widehat{R}$ \citep{gelm92}. An adapted version of $\widehat{R}$ has been proposed by \cite{veht19}, which may also assess the stationarity component of convergence. Therefore, the focus of this study is mainly on $\widehat{R}$. Stationarity will also be evaluated with auto-correlation as diagnostic, as recommended by \cite[p.~898]{cowl96}. Other convergence diagnostics are outside the scope of this study. \footnotemark 
\footnotetext{All of these methods evaluate the convergence of univariate scalar summaries (e.g., chain means or variances). These convergence diagnostics cannot diagnose convergence of multivariable statistics (i.e., relations between scalar summaries). \cite{buur18} proposed to implement multivariable evaluation by means of eigenvalue decomposition \cite{mack03}. This method is outside of the scope of the current study.}%  he eigenvector decomposition method proposed by McKay \citep{XYZ}.} These four methods, however, are outside of the scope of this study. Here, the focus is mainly on $\widehat{R}$ and auto-correlation.

%Diagnose non-convergence in the mixing sense: $\widehat{R}$. Diagnose non-convergence in the stability sense: auto-correlation (AC). Also, it is possible to compute MC errors, the Geweke statistic \textbf{\citep{XYZ}}, or to determine the appropriate number of iterations for convergence by employing \textbf{Raftery and Lewis's (1992)} method. All of these methods evaluate the convergence of univariate scalars of interest (e.g., chain mean or chain variance). The convergence of multivariable statistics (i.e. relations between such scalars) may be evaluated via the eigenvector decomposition method proposed by McKay \citep{XYZ}. These four methods, however, are outside of the scope of this study. Here, the focus is mainly on $\widehat{R}$ and auto-correlation.}

It is not known whether any convergence diagnostic is appropriate to diagnose convergence of the MICE algorithm. As recommended by \cite[\S~6.5.2]{buur18}, the current practice is to visually inspect imputation chains for signs of non-mixing and non-stationarity. Not only is visually assessing convergence challenging to the untrained eye, it might also be futile. That is, there is no scientific consensus on the convergence properties of multiple imputation algorithms \citep{taka17}, and in some situations the default techniques in \pkg{mice} might not converge to stable target distributions at all \citep{murr18}. %Convergence diagnostics for MI methods have not been systematically studied \citep{buur18}.  Moreover, 
%A fundamental assumption of MICE is convergence of the algorithm. MICE is a type of Markov chain Monte Carlo (MCMC) algorithm. In general, the validity of inference resulting from MCMC algorithms is threatened by non-convergence. Hence we use convergence diagnostics to flag non-convergence. But it is not known whether conventional convergence diagnostics for MCMC methods work on MI data.
Therefore, the need for convergence diagnostics is extra pressing in the context of the MICE algorithm.

@@

@@

%We can apply $\widehat{R}$ to the mean (or to the first two moments) of the variables of interest. 
$\widehat{R}$ is defined by \cite{veht19} as follows: % The scalar summary of interest in this study is the average of the imputed values per variable. %As recommended by \cite{veht19} $\widehat{R}$ is computed as .... to be able to detect ... in the tails of the distribution. \cite[p.~5]{veth19}

@@

The second component of convergence stationarity requires chains to be stable across iterations, or non-recurring. Non-recurrence can be evaluated with auto-correlation. Auto-correlation shows how dependent subsequent draws of an imputation chain are on the previous value. If there is a lot of dependence, draws at e.g. iteration five are significantly correlated with the value of the first draw. A high auto-correlation indicates dependence within chains. The magnitude of the AC can be interpreted qualitatively, or compared to the critical value of a \textbf{XYZ} distribution with \textbf{XYZ} degrees of freedom.

\textbf{Convergence diagnostics applied to MI}

@@

%Conventional thresholds to diagnose non-convergence-- e.g., Gelman and Rubin's \citeyear{gelm92} statistic $\widehat{R} < 1.1$ --are not applicable on multiply imputed data \citep{lace07} \textbf{[Expand or remove Rhat here! Refer to future section?]}. Therefore, empirical researchers have to rely on visual inspection procedures that are theoretically equivalent to $\widehat{R}$ \citep{whit11} \textbf{[explain why this can still be appropriate, and numerical is not]}. Visually assessing convergence is not only challenging to the untrained eye, it might also be futile.

@@

\subsection{Data simulation}
The simulated dataset is a finite population of $N=1000$. The data are simulated to solve a multiple linear regression complete data problem. The quantity of scientific interest is the estimated regression coefficient of predictor $X$ on outcome variable $Y$. The linear regression model is $Y \sim \beta_1 X + \beta_2 Z_1 + \beta_3 Z_2$, where $Y$ is the dependent variable, $X$ is an independent variable, and $Z_1$ and $Z_2$ are covariates. The data generating model of the predictors is a multivariate normal distribution with means structure $\mu$, and variance-covariance matrix $\Sigma$.

@@

\subsection{Simulation diagnostics}
The simulation diagnostics are as recommended by \cite[\S~2.5.2]{buur18}. These evaluation criteria comprise of average bias, average confidence interval width, and empirical coverage rate across simulations. %"We expect that in accordance with Neyman (1934), at least 95 percent of the confidence intervals should contain the true population value, although some room for simulation error should be taken into account to counteract the finite nature of our simulations"\cite{schou18}.
We could also look at distributional characteristics, and plausibility of imputed values, see \cite{vinknd} (n.d.). For now, this is outside of the scope of this study. % Or use: Then evaluate the regular diagnostics of MI simulations, see Vink, n.d., and  convergence. Diagnostics include absolute bias of the estimated regression coefficient, confidence interval width, and empirical coverage rate across simulations. Convergence is evaluated with $\widehat{R}$, or potential scale reduction factor.

@@

Figures \ref{fig:conv} and \ref{fig:sim} display results of the simulation study aggregated over 1000 repetitions. A subset of simulation conditions is presented in Table \ref{tab:results}. This section describes convergence diagnostics ($\widehat{R}$ and auto-correlation) and simulation diagnostics (bias, confidence interval width and coverage rate) first separately, and then together.

@@

\textbf{$\widehat{R}$}. The $\widehat{R}$ values for chain means and chain variances show equivalent trends, neither is there a clear difference between the four variables. Roughly speaking, we see that $\widehat{R}$ is higher for simulation conditions with a lower number of iterations, and approaches one for conditions with more iterations. Figure \ref{fig:conv} shows an initial dip, a small bump upwards, and then a steep decline. This decline \textbf{houdt aan} up-to iteration twenty, after which a more gradual decrease can be observed up-to iteration forty. The $\widehat{R}$ values are more or less stable after that. The conventional acceptable threshold for convergence, maximum $\widehat{R} < 1.2$, is reached in the simulation condition with three iterations, and conditions with six iterations or more. The widely used threshold $\widehat{R} < 1.1$ is met in conditions with ten or more iterations. The most recent recommended threshold is not reached within the 100 iterations considered in this simulation study. $\widehat{R} > 1.01$ for all simulation conditions (see online appendix for the full table of results). \textbf{Check is computing r ht is still correct!}

\textbf{Auto-correlation.} All observed auto-correlations (aggregated over the 1000 repetitions) are negative or close to zero. There are some differences between variables, and the ACs of chain variances seem lower than the ACs of chain means. The general trend in figure \ref{fig:conv} is that ACs decrease steeply as the number of iterations increase.  \textbf{The lowest AC observed is %\Sexpr{round(min(dat[-1, c("AC chain mean", "AC chain var")]), 3)} 
at four iterations. Simulation conditions with a higher number of iterations have show a slow increase towards zero. Auto-correlations do not plateau completely within the range of simulation conditions considered in this study. It seems that ACs might become positive if more iterations would be allowed. The average absolute maximum ACs are negative up to 77 iterations for chain means, and for chain variances all iterations considered here have negative ACs.}

@@

%% --------------------------------
\subsection{Simulation diagnostics}
\textbf{Bias.} Figure \ref{fig:sim} shows the average bias between the regression coefficient of $X$ on $Y$ in the complete dataset, and $\bar{Q}$, \textbf{the pooled estimated regression coefficient and the  is close to zero for simulation conditions with two iterations or more. The simulation condition with one iteration has a negative bias. From two iterations upwards, the average bias across repetitions is stable. Across iterations two to one hundred, bias fluctuates within a narrow range around zero. Loess line is flat around it = 20.} 
% Average bias is not zero because there is only one incomplete dataset to be imputed. Bias is a sample effect.

\textbf{Confidence interval width.} CIW is only clearly divergent for the simulation condition with one iteration. Conditions with two or more iterations have similar CIWs. The magnitude of differences from the loess line is somewhat larger up-to the condition with four iterations. From four iterations on, CIW seems to be stable across iterations. \textbf{Loess line is never flat.}%One might argue that there is some downward trending up-to thirthy to forty iterations. %After about twenty-five iterations the average CIW across all simulation conditions and repetitions is reached \Sexpr{round(mean(dat[, "CI width"]), 3)}.

\textbf{Coverage rate.} \textbf{The coverage rate is more or less stable from two iterations upwards. On average, the coverage rate is somewhat higher than the expected nominal coverage of 95\% \cite{neym34}, namely \Sexpr{100*round(mean(dat[, "Cov. rate"]), 3)}\%. Loess line is never flat.}

\textbf{give some implications for these measures. connect convergence and simulation diagnostics.}

@@

\textbf{Answer RQ.} How to diagnose non-convergence? It's complicated.

@@

\textbf{Future research.} %We would like to have convergence measures for multivariable statistics (scalars?) of interest. This is, however, dependent on the complete data model. The eigenvector decomposition method proposed by McKay (?) should be implemented. I could not find any resources to apply this method and it is outside the scope of this thesis to investigate how this approach could be implemented. \textbf{[this paragraph is too negative, explain scope in intro instead.]}

@@

### 2019-12-11 Simulation/Writeup/NoteConvergence.tex
ran full simulation with convergence diagnostics per var instaed of max

\textbf{$\widehat{R}$}. Steep decline up-to iteration twenty, then more gradual decrease up-to iteration forty, more or less stable after that. Conventional threshold $\widehat{R} < 1.2$ reached after seven or eight iterations, $\widehat{R} < 1.1$ after fourteen iterations for chain means and sixteen for chain variances, $\widehat{R} < 1.01$ not reached within the 100 iterations considered in this simulation study (see online appendix for the full table of results). \textbf{[REWRITE AS STORY, NOT JUST DATA.]}

\textbf{Auto-correlation.} Within the 100 iterations considered in this study, the AC decreases steeply, before a slow increase. The lowest AC observed is -0.735 at four iterations. There appears to be a trend towards zero from four iterations upwards. Auto-correlation does not plateau completely. The average absolute maximum ACs are negative up to 77 iterations for chain means, and for chain variances all iterations considered here have negative ACs.

@@

%% --------------------------------
\subsection{Simulation diagnostics}
\textbf{Bias.} From two iterations upwards, the average bias across repetitions is stable. Across iterations two to one hundred, bias fluctuates within a narrow range around zero. 
% Average bias is not zero because there is only one incomplete dataset to be imputed. Bias is a sample effect.

\textbf{Confidence interval width.} CIW does not show a clear trend across iterations. One might argue that there is some downward trending up-to thirthy to forty iterations. %After about twenty-five iterations the average CIW across all simulation conditions and repetitions is reached 0.928.

\textbf{Coverage rate.} The coverage rate is more or less stable from two iterations upwards. On average, the coverage rate is somewhat higher than the expected nominal coverage of 95\% \cite{neym34}, namely 95.3\%.

\textbf{add loess line to plots to see when it becomes a flat line. also, give some implications for these measures.}

@@

\textbf{LEAVING THIS HERE FOR REFERENCE, PLEASE REMOVE BEFORE HANDING IN}

\includegraphics{NoteConvergence-003}

>>>>

\textbf{$\widehat{R}$}. The $\widehat{R}$ values for chain means and chain variances show equivalent trends, neither is there a clear difference between the four variables. Roughly speaking, we see that $\widehat{R}$ is higher for simulation conditions with a lower number of iterations, and approaches one for conditions with more iterations. Figure \ref{fig:conv} shows an initial dip, a small bump upwards, and then a steep decline. This decline \textbf{houdt aan} up-to iteration twenty, after which a more gradual decrease can be observed up-to iteration forty. The $\widehat{R}$ values are more or less stable after that. The conventional acceptable threshold for convergence, maximum $\widehat{R} < 1.2$, is reached in the simulation condition with three iterations, and conditions with six iterations or more. The widely used threshold $\widehat{R} < 1.1$ is met in conditions with ten or more iterations. The most recent recommended threshold is not reached within the 100 iterations considered in this simulation study. $\widehat{R} > 1.01$ for all simulation conditions (see online appendix for the full table of results). \textbf{Check is computing r ht is still correct!}

\textbf{Auto-correlation.} All observed auto-correlations (aggregated over the 1000 repetitions) are negative or close to zero. There are some differences between variables, and the ACs of chain variances seem lower than the ACs of chain means. The general trend in figure \ref{fig:conv} is that ACs decrease steeply as the number of iterations increase.  \textbf{The lowest AC observed is %-0.735 
at four iterations. Simulation conditions with a higher number of iterations have show a slow increase towards zero. Auto-correlations do not plateau completely within the range of simulation conditions considered in this study. It seems that ACs might become positive if more iterations would be allowed. The average absolute maximum ACs are negative up to 77 iterations for chain means, and for chain variances all iterations considered here have negative ACs.}

@@

%% --------------------------------
\subsection{Simulation diagnostics}
\textbf{Bias.} Figure \ref{fig:sim} shows the average bias between the regression coefficient of $X$ on $Y$ in the complete dataset, and $\bar{Q}$, \textbf{the pooled estimated regression coefficient and the  is close to zero for simulation conditions with two iterations or more. The simulation condition with one iteration has a negative bias. From two iterations upwards, the average bias across repetitions is stable. Across iterations two to one hundred, bias fluctuates within a narrow range around zero. Loess line is flat around it = 20.} 
% Average bias is not zero because there is only one incomplete dataset to be imputed. Bias is a sample effect.

\textbf{Confidence interval width.} CIW is only clearly divergent for the simulation condition with one iteration. Conditions with two or more iterations have similar CIWs. The magnitude of differences from the loess line is somewhat larger up-to the condition with four iterations. From four iterations on, CIW seems to be stable across iterations. \textbf{Loess line is never flat.}%One might argue that there is some downward trending up-to thirthy to forty iterations. %After about twenty-five iterations the average CIW across all simulation conditions and repetitions is reached 0.928.

\textbf{Coverage rate.} \textbf{The coverage rate is more or less stable from two iterations upwards. On average, the coverage rate is somewhat higher than the expected nominal coverage of 95\% \cite{neym34}, namely 95.3\%. Loess line is never flat.}

\textbf{give some implications for these measures. connect convergence and simulation diagnostics.}

@@
